Characterizing Signal Loss, Error, and Bias in the 21cm Reionization \\Power Spectrum: A Revised Study of PAPER-64

### 2018-03-23 PAPER.tex
Polished up appendix

In this Appendix, we examine a toy model for signal loss. Our goal is to derive an analytic formula for power spectrum signal loss. We will also show that in general, signal loss appears as a multiplicative bias on one's power spectrum estimate.

@@

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\widehat{p}_\alpha^\textrm{loss}$ in terms of $\widehat{p}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\widehat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\widehat{p}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading 
order in $\eta$. This approximates the limiting case where the covariance $\Chat$ is close to the ideal and the 
lossy covariance is a small perturbation.  The result is
\begin{equation}
\widehat{p}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \widehat{p}_\alpha^\textrm{loss} \rangle \approx (1- \eta N) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N) p_\alpha,
\end{equation}
where $N$ is the length of $\x$. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained. Now, recall that $p_\alpha$ is the \emph{true} power spectrum. This means that it can be decomposed into the sum of the foreground and EoR power spectra, since the foregrounds and EoR are uncorrelated in expectation. This toy example, while not showing the emergence of cross-terms provides a rudimentary motivation for the form of Equation \eqref{eq:LinearPspecSum}.

>>>>

In this Appendix, we examine a toy model for signal loss. Our goal is to derive an analytic formula for power spectrum signal loss. While this 
model does not apply generally to all the scenarios presented in this paper, it provides some analytic intuition for how the coupling between data and an empirical covariance can result in signal loss.

@@

Our strategy for computing the signal loss will be to insert Equation \eqref{eq:ChatDef} into Equation \eqref{eq:phatloss} and to express the resulting estimator $\widehat{p}_\alpha^\textrm{loss}$ in terms of $\widehat{p}_\alpha$. We begin by expressing $\Chat^{-1}$ in terms of $\C^{-1}$ using the Woodbury identity so that
\begin{equation}
\Chat^{-1} = \frac{\C^{-1}}{1-\eta} \left[ \I - \frac{\eta \x \x^t \C^{-1}}{1+ \eta (g-1)}\right],
\end{equation}
where we have defined $g \equiv \x^t \C^{-1} \x$. Inserting this into our Fisher estimate we have
\begin{equation}
\widehat{F}_{\alpha \alpha} = \frac{F_{\alpha \alpha}}{(1-\eta)^2} \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right],
\end{equation}
where $h_\alpha \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \x $ and $h_{\alpha \alpha} \equiv \x^t \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1}\x $. Note that $g$, $h_\alpha$, and $h_{\alpha \alpha}$ are all random variables, since they depend on $\x$. Inserting these expressions into our estimator gives
\begin{equation}
\label{eq:phatlossexpanded}
\widehat{p}_\alpha^\textrm{loss} = \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \frac{\eta g}{1+ \eta (g-1)}\right]^2  \left[ 1 -\frac{\eta }{1+ \eta (g-1)} \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}} + \frac{1}{2} \left( \frac{\eta }{1+ \eta (g-1)} \right)^2 \frac{h_\alpha^2}{F_{\alpha \alpha}}\right]^{-1}.
\end{equation}
Both for the purposes of analytical tractability and to provide intuition, we expand this expression to leading 
order in $\eta$. This approximates the limiting case where the covariance $\Chat$ is close to the ideal and the 
lossy covariance is a small perturbation.  The result is
\begin{equation}
\widehat{p}_\alpha^\textrm{loss} \approx \frac{1}{2} \frac{h_\alpha}{F_{\alpha \alpha}} \left[ 1 - \eta \left( g - \frac{h_{\alpha \alpha}}{F_{\alpha \alpha}}\right)\right].
\end{equation}
Taking the ensemble average of both sides and noting that the true power spectrum $p_\alpha$ is equal to $\langle h_\alpha \rangle / 2 F_{\alpha \alpha}$, we obtain
\begin{equation}
\langle \widehat{p}_\alpha^\textrm{loss} \rangle \approx (1- \eta N) p_\alpha + 4 \eta \frac{\rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha \C^{-1} \Q^\alpha )}{\left[ \rm{tr} (\C^{-1} \Q^\alpha \C^{-1} \Q^\alpha  ) \right]^2} \approx (1- \eta N) p_\alpha,
\end{equation}
where $N$ is the length of $\x$. In the last step we dropped the final term, since it scales as $\eta p_\alpha$ (without the factor of $N$) and is therefore typically small compared to the terms that have been retained.

Recalling that $p_\alpha$ is the \emph{true} power spectrum, one sees that when the covariance in the optimal quadratic estimator is naively replaced by an empirical covariance, the resulting power spectrum estimate is biased low, i.e., there is signal loss. This occurs because of couplings between $\widehat{\C}$ and $\x$, which formally means that what was original a quadratic estimator is no longer quadratic, but contains higher-order correlations. This violates the assumptions implicit in the derivation of $F_{\alpha \alpha}$ as the normalization factor for converting unnormalized bandpowers $\frac{1}{2} \x^t \C^{-1} \Q^{\alpha} \C^{-1} \x$ into properly normalized power spectrum estimates, where the unnormalized bandpowers are assumed to be two-point (i.e., quadratic) statistics \citep{LT11}. The result is an improperly normalized---and thus lossy---power spectrum estimate.
